---
title: "Taste Influence Analysis: Output Generation"
author: "Omar Lizardo & Sara Skiles"
date: "June 2026"
format:
  html:
    toc: true
    theme: flatly
    code-fold: true
    embed-resources: true
---

# Analysis Pipeline

This reproducible document executes the models and generates the tables and figures used in the main manuscript.

## 1. Data Setup & Propensity Score Weighting

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(dplyr)
library(ggplot2)
library(modelsummary)
library(marginaleffects)
library(ordinal)
library(nnet)
library(WeightIt)
library(cobalt)

# Execute data processing
source("R/dataproc.R")
data_list <- process_data()
df_long <- data_list$long
df_wide <- data_list$wide

# Generate Complete Cases for Weighting
df_wide_cc <- df_wide %>%
  filter(!is.na(objsubjclass_factor), 
         !is.na(age), !is.na(female), !is.na(raceeth), !is.na(parented))

# Compute Inverse Probability Weights (IPW)
W <- weightit(objsubjclass_factor ~ age + female + factor(raceeth) + parented, 
              data = df_wide_cc, method = "ps", estimand = "ATE")

df_wide_cc$ipw_weight <- W$weights

# Join weights back to the long dataset for mixed models
df_long <- df_long %>%
  inner_join(df_wide_cc %>% select(id, ipw_weight), by = "id")
```

## 2. Covariate Balance Check

```{r balance, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
p_balance <- love.plot(W, binary = "std", thresholds = c(m = .1), 
          title = "Covariate Balance Before and After Weighting") +
  theme_minimal()
print(p_balance)
ggsave("figures/Figure_CovariateBalance.png", p_balance, width = 8, height = 6)
```

## 3. Linear Baseline and Trial Effects (Weighted)

```{r table2, warning=FALSE, message=FALSE}
mod0 <- lmer(taste ~ 1 + (1 | id), data = df_long, weights = ipw_weight, REML = FALSE)
mod1 <- lmer(taste ~ cond_factor + (1 | id), data = df_long, weights = ipw_weight, REML = FALSE)
mod2 <- lmer(taste ~ factor(trial) + cond_factor + (1 | id), data = df_long, weights = ipw_weight, REML = FALSE)
mod3 <- lmer(taste ~ factor(trial) * cond_factor + (1 | id), data = df_long, weights = ipw_weight, REML = FALSE)
anova(mod0, mod1, mod2, mod3)
```

## 4. Trial Effects by Status Consistency (Weighted)

```{r fig5, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
mod_objsubj <- lmer(taste ~ factor(trial) * cond2_factor * objsubjclass_factor + (1 | id), 
                    data = df_long, weights = ipw_weight, REML = FALSE)
te_objsubj <- avg_comparisons(mod_objsubj, variables = "trial", by = c("cond2_factor", "objsubjclass_factor"))

te_objsubj <- te_objsubj %>%
  mutate(
    cond_clean = factor(cond2_factor, levels = c(
      "Class & Taste/Dislike/-Status",
      "Class & Taste/Dislike/+Status",
      "Taste Only/Dislike",
      "Baseline",
      "Taste Only/Like",
      "Class & Taste/Like/-Status",
      "Class & Taste/Like/+Status"
    ), labels = c(
      "Dislike / Low Status",
      "Dislike / High Status",
      "Dislike (No Status)",
      "Baseline (No Info)",
      "Like (No Status)",
      "Like / Low Status",
      "Like / High Status"
    )),
    sig = ifelse(conf.low > 0 | conf.high < 0, "Significant", "Not Significant")
  )

p_forest <- ggplot(te_objsubj, aes(x = estimate, y = cond_clean, color = sig)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2, linewidth = 1, orientation = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray") +
  facet_wrap(~objsubjclass_factor, ncol = 2) +
  scale_color_manual(values = c("Significant" = "firebrick", "Not Significant" = "grey50")) +
  labs(title = "Estimated Trial Effect by Condition and Status (IPW Weighted)",
       x = "Shift in Evaluation (Taste 2 - Taste 1)", y = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", legend.title = element_blank())

print(p_forest)
ggsave("figures/Figure5_TrialEffects_Forest.png", p_forest, width = 10, height = 8)
```

## 5. Cumulative Link Mixed Models (CLMM - Weighted)

```{r clmm, warning=FALSE, message=FALSE, eval=FALSE}
df_long$taste_ord <- factor(df_long$taste, ordered = TRUE, levels = 1:7)
clmm_mod <- clmm(taste_ord ~ factor(trial) * cond2_factor + (1 | id), 
                 data = df_long, weights = ipw_weight)

# Display interaction coefficients
clmm_coefs <- coef(clmm_mod)[grep("trial.*cond", names(coef(clmm_mod)))]
round(clmm_coefs, 3)
```

## 6. Disaggregating Behavior: Conformity vs. Reactance (Weighted)

```{r multinom, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
df_multi <- df_wide_cc %>%
  mutate(
    diff = taste2_rev - taste1_rev,
    behavior = case_when(
      cond2_factor == "Baseline" & diff == 0 ~ "Stay",
      cond2_factor == "Baseline" & diff > 0 ~ "Shift Up",
      cond2_factor == "Baseline" & diff < 0 ~ "Shift Down",
      cond2_factor %in% c("Class & Taste/Like/-Status", "Class & Taste/Like/+Status", "Taste Only/Like") & diff == 0 ~ "Stay",
      cond2_factor %in% c("Class & Taste/Like/-Status", "Class & Taste/Like/+Status", "Taste Only/Like") & diff > 0 ~ "Conform",
      cond2_factor %in% c("Class & Taste/Like/-Status", "Class & Taste/Like/+Status", "Taste Only/Like") & diff < 0 ~ "React",
      cond2_factor %in% c("Class & Taste/Dislike/-Status", "Class & Taste/Dislike/+Status", "Taste Only/Dislike") & diff == 0 ~ "Stay",
      cond2_factor %in% c("Class & Taste/Dislike/-Status", "Class & Taste/Dislike/+Status", "Taste Only/Dislike") & diff < 0 ~ "Conform",
      cond2_factor %in% c("Class & Taste/Dislike/-Status", "Class & Taste/Dislike/+Status", "Taste Only/Dislike") & diff > 0 ~ "React",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(behavior %in% c("Stay", "Conform", "React")) %>%
  mutate(behavior = factor(behavior, levels = c("Stay", "Conform", "React")))

multi_mod <- multinom(behavior ~ objsubjclass_factor, 
                      data = df_multi, weights = ipw_weight, trace = FALSE)
multi_preds <- predictions(multi_mod, type = "probs", by = c("objsubjclass_factor", "group"))

p_multi <- ggplot(multi_preds, aes(x = estimate, y = objsubjclass_factor, color = group)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), position = position_dodge(width = 0.5), width = 0.2, linewidth = 1, orientation = "y") +
  scale_color_manual(values = c("Stay" = "forestgreen", "Conform" = "steelblue", "React" = "firebrick")) +
  labs(title = "Probability of Behavioral Response by Status Group (IPW Weighted)",
       x = "Predicted Probability", y = "", color = "Behavior") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

print(p_multi)
ggsave("figures/Figure7_MultinomialBehavior.png", p_multi, width = 10, height = 6)
```

## 7. Sensitivity Analysis: Isolating Class Feedback (Weighted)

```{r sensitivity, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
# Filter out "Taste Only" conditions
df_multi_sens <- df_multi %>%
  filter(!cond2_factor %in% c("Taste Only/Like", "Taste Only/Dislike")) %>%
  mutate(cond2_factor = droplevels(cond2_factor))

multi_mod_sens <- multinom(behavior ~ objsubjclass_factor, 
                           data = df_multi_sens, weights = ipw_weight, trace = FALSE)

multi_preds_sens <- predictions(multi_mod_sens, type = "probs", by = c("objsubjclass_factor", "group"))

p_sens <- ggplot(multi_preds_sens, aes(x = estimate, y = objsubjclass_factor, color = group)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), position = position_dodge(width = 0.5), width = 0.2, linewidth = 1, orientation = "y") +
  scale_color_manual(values = c("Stay" = "forestgreen", "Conform" = "steelblue", "React" = "firebrick")) +
  labs(title = "Behavioral Response (Excluding 'Taste Only' Conditions)",
       x = "Predicted Probability", y = "", color = "Behavior") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

print(p_sens)
ggsave("figures/Figure11_Sens_MultinomialBehavior.png", p_sens, width = 10, height = 6)
```
